Use of general purpose graphical processing units with MODFLOW

### 2012-09-04 ResponseTo_GW20120518-0099_InternalReview.tex

\noindent
\textbf{Use of general purpose graphical processing units with MODFLOW-2005 \\ (Ground Water Manuscript GW20120518-0099) }\\

\noindent
\textbf{Authors:} Hughes, J.D., and White, J.T \\

\today \\

\noindent
Internal reviews were performed by (1) Dr. Yong Liu a senior research scientist at the National Center for Supercomputing Applications (NCSA) University of Illinois at Urbana-Champaign, (2) Randy Hanson a research hydrologist in the California Water Science Center, and (3) Vevek Bedekar a professional engineer with  S.S. Papadopulos \& Associates, Inc. Randy Hanson provided a few comments in a review memorandum, which are included and addressed below, and as inline comments in the manuscript pdf; inline comments are addressed in the document titled MODFLOW-GPU\_03.jdhResponse.pdf. Internal review comments were helpful and have been addressed in the revised manuscript. Review comments are included (\textit{in italics}) prior to our responses. Revised material in the manuscript are indicated with \color{cyan} cyan \color{black} text if it has not been addressed previously in the responses to journal reviewer comments (which are indicated with \color{blue} blue \color{black} text).

@@

\section*{\small{Yong Liu Comments:}}
\begin{enumerate}
\item \textit{One thing that I would suggest to add is to say a few words about any possible future research (e.g., GPU cluster, Cloud-based GPU resources such as Amazon GPU cluster) and practical usage of GPU solvers for MODFLOW 2005 for real-world use cases (an example real-world use case that would benefit from using this research findings and the GPU solver would be excellent!.}

We have added a brief discussion of current research being conducted by others of improved parallelization of incomplete factorization preconditioners and there is a discussion of possible future enhancements of the UPCG solver in the last pragraph of the \textbf{Conclusions} section.

We have not added a practical example because we believe the test problems presented represent a synthetic but realistic system and is a good control problem for evaluating the effectiveness of the UPCG solver -- and this is the ultimate goal of this paper.

\item \textit{It would be also nice if the authors can indicate where and how to obtain the source code (will it be in public domain just like the MODFLOW source code?).}

We have included a new section titled \textbf{How to Obtain the Software} that includes this information.

@@

\item \textit{It would be nice if you could also provide the python code so that I or someone else can reproduce some of your experiments.}

This will be included in the github repository that is described in the \textbf{How to Obtain the Software} section.

@@

\end{enumerate}

\section*{\small{Randy Hanson Comments:}}
\begin{enumerate}
\item \textit{In particular the preconditioning seems to be an issue and one is left wondering if the code could be written better or is it simply a hardware issue given the structure of the code. I have made several inline comments that may give more details as to where these issues might require minor additional elaboration or insight based on your experience. Would a graph showing the change in performance while varying the inner and outer iterations for one of the models provide any additional insight on how to best set the combination of these two, or is this too problem dependent?}

Undoubtedly, there is room for improving the efficiency of the UPCG solver code (and probably any experimental code) but we believe the issues related to parallelization of the Jacobi, GLSPOLY, and MILU0 preconditioners would still apply even if the code were more efficient. However having said that, we believe there is some problem dependence to the optimal number of Picard and inner iterations for a given problem. We set up the test problem so that we maximized the amount of time spent in the linear solver to maximize the conjugates evaluated and allowed a large number of Picard iterations to ensure convergence was achieved for most simulations.

See responses to inline comments in MODFLOW-GPU\_03.jdhResponse.pdf.

@@

\item \textit{The reader should also have a clear idea of how he could implement these choices and which choice is best for the problem at hand to some degree (although this may be outside the scope of this paper).}

We have included our best attempt at guidance based on our simulations for effectively using the various preconditioners and/or hardware options in the \textbf{Discussion} section.

@@

\item \textit{The reader may also wonder if additional compilation features such as provided by Intel compiler for memory alignment, vectorization, use of coarrays, use of math-kernal library vectorized functions, or chip specific architectures could provide additional performance enhancements.}

We haven't seen significant improvements with chip specific compiler switches, MKL libraries (beyond improvements related to OpenMP), data alignment, or vectorization (the UPCG solve is a vector solver that uses CRS indirect indexing). Co-arrays may be beneficial for cluster applications but probably not as beneficial for multi--core CPUs because of the limited number of available threads. A cluster solution is something that we are considering but would be less accessible to many readers of Ground Water. We suspect that parallel CPU and GPGPU solutions would be more accessible to readers.

@@

\section*{Vevek Bedekar Comments:}
\begin{enumerate}
\item \textit{Specify whether this work is first of its kind in applying GPUs in the field of groundwater modeling or specifically in the application of MODFLOW. Otherwise, cite other works that apply GPUs to solve groundwater simulations. The following two search results were found as a result of a quick Google search.}

\textit{\href{http://doi.acm.org/10.1145/1774088.1774588}{Ji, Xiaohui and Cheng, Tangpei and Wang, Qun, 2010, A simulation of large-scale groundwater flow on CUDA-enabled GPUs: Proceedings of the 2010 ACM Symposium on Applied Computing, 2402--2403. doi: 10.1145/1774088.1774588}}

\textit{\href{http://www.gpucomputing.net/?q=node/14182}{Ji, Xiaohui and Cheng, Tangpei and Wang, Qun, 2010, CUDA-based solver for large-scale groundwater flow simulation: Engineering with Computers, 28(1), 13--19. doi: 10.1007/s00366-011-0213-2}}

The second reference (Ji \textit{et al.}, 2012) is a reference in the manuscript but this is difficult to determine from the link provided. The first reference (Ji \textit{et al.}, 2010) was not included in the original manuscript since it is a conference proceeding and may be difficult for some readers to locate. It has been added to ensure Ji \textit{et al.}, 2010 is credited as being the first application of solving MODFLOW using GPGPUs.

\item \textit{Define the term speedup in the paper. Speedup is used in many different ways: solver speedup, speedup of parallelized loops, or speedup of total simulation time. Since simulation efficiency is at the core of this paper, it will be useful to mention what the calculated speedups are based on.}

The definition of speedup has been clarified to indicate that it is based solely on the time spent within solver routines.

\item \textit{Are there any drawbacks to using GPUs and how these compare to using CPUs: limitations on the size of the problem; a general cost comparison (twice/thrice) etc.}

Information on the RAM limitations of the specific GPGPUs is given in the \textbf{Test Cases} and \textbf{Conclusions} sections. A summary under which use of parallel GPGPU and CPU approaches should be beneficial has been added to the \textbf{Discussion} section.

\item \textit{Page 9 mentions that two different machines were used (2.4 GHz and 3 GHz) for comparing the CPU and GPU simulations. Were comparisons made between the two machines by running exactly same simulation runs on the two CPUs (without multithreading and by using 2 cores) to benchmark whether there are inherent differences in the performance of the two machines? Besides the processor speed other factors like bus speed, motherboard, hard drives, etc. could be impacting the results.}

Serial and parallel runs were made on both machines so that the correct speedup could be calculated on each machine. Additional text has been added to clarify the approach used for model simulations on both machines.

\item \textit{Did the authors study if a relationship existed between the total number of inner iterations and the speedups in the test simulations? If such a comparison was made, it may be interesting to share that with the readers. That would be an indication to the users as to which problems may reap the benefits of GPU more than others.}

Yes there is a relationship between inner iterations and speedup. The runtime information shown in \textbf{figure 4} tells part of the story but marginal to poor speedups for the Jacobi preconditioner is related to the number of iterations required for convergence.

A new table has been added (\textbf{table 1} in the revised manuscript) that summarizes the number of Picard and inner iterations used by each solver and preconditioner evaluated. Additional text that discusses the relationship between the strength of the preconditioner and the observed speedup has been added to the \textbf{Discussion} section. We believe this will provide readers with some guidance on when to try parallel GPGPU or CPU approaches to a given problem.

\item \textit{Was the total number of inner and outer iterations compared between the CPU and GPU simulations? The number should come out to be very similar in all simulations since the only thing GPU is doing is to parallelize the same computations that CPU is doing. This comparison would provide more confidence in the results that no other processes/algorithms besides GPU and its parallelization are responsible for the speedups.}

A new table has been added (\textbf{table 1} in the revised manuscript) that summarizes the number of Picard and inner iterations used by each solver and preconditioner evaluated. Additional text has been added to the \textbf{Results} section to summarize the data included on the new table.

\item \textit{There are more robust and faster solvers available besides PCG2 (XMD of MODFLOW‐NWT; PCG5 of MODFLOW‐SURFACT). It would be interesting to see how the GPU speedup (of 8 when compared to PCG2), compares with faster algorithms used in other solvers like XMD and PCG5.}

There are certainly other solvers available that have been incorporated into MODFLOW or MODFLOW-variants by others but we have intentionally limited our discussion to standard solvers available for MODFLOW-2005. This study was meant to be an evaluation of the applicability of GPGPU parallelization to a PCG solver not all solvers. The findings of this study should be applicable to other solvers and could be implemented in a similar fashion. It is our expectation that speedups for the solvers mentioned would be similar to the UPCG solver with the MILU0 preconditioner because of the difficulty of parallelizing an incomplete factorization.

It is our assessment that it is the parallel GLSPOLY preconditioner that makes use of the GPGPU useful for MODFLOW at this point. Furthermore, we expect that future research on ways to parallelize incomplete factorization preconditioners will greatly improve the speedup for parallel solvers on CPUs and GPGPUs.

\item \textit{The LMG2 (SAMG) solver is a parallel solver that claims speedups of up to 100 times when compared to the PCG solver  (\url{http://www.scai.fraunhofer.de/fileadmin/download/samg/paper/modflow11.pdf}; \url{http://www.scai.fraunhofer.de/fileadmin/download/samg/paper/Modflow_Paper.pdf}). The authors may wish to comment about the combination of the GPU technology and the SAMG
solver.}

The authors have considered parallelization of a multi-grid solver but have not performed any evaluation of the effectiveness of parallelization of a multi-grid solver relative to a traditional Krylov solver. Parallel multi-grid solvers are reported to be more scalable than traditional Krylov solvers. To keep the paper on topic - comparison of parallel strategies on the UPCG solver with speedups based on the PCG solver - we have elected not to add a discussion of parallelization of the LMG2 solver on GPGPUs.

\item \textit{How does PCG and UPCG sequential simulation compare?}

PCG and UPCG are competitive. For some problems PCG is slightly faster for other problem UPCG is slightly faster. Iteration information is summarized on \textbf{table 1} in the revised manuscript. In general, the performance of PCG suffers with larger problems and is summarized in the \textbf{Results} section.

\item \textit{Mention that the groundwater simulation results are the same irrespective of the solver/processor used.}

This has been added to the end of the \textbf{Problem Description} subsection.

\item \textit{Line 35: replace SOR with SSOR.}

Corrected

\item \textit{Line 40: also refer to other more efficient solvers, including the XMD, LMG2, and PCG5 solvers.}

There are certainly other solvers available that have been incorporated into MODFLOW or MODFLOW-variants by others but we have intentionally limited our discussion to standard solvers available for MODFLOW-2005. XMD has been incorporated into MODFLOW-NWT but it's real strength is its ability to solve non-symmetric matrices using more sophisticated level-fill and drop tolerance approaches. Since the test problems we are evaluating are symmetric it is questionable whether these approaches would add convergence improvements that outweigh the additional numerical overhead of these approaches. In my experience XMD is less efficient in simulating standard MODFLOW problems than the PCG and UPCG solvers.

\item \textit{Line 44: it is important to note that speedup in this paper is based on ``solver computing time'' as defined in the paper. Upon testing, I had found that the loops parallelized took 25--30\% of the total simulation time for the problem I was testing. Effectively, 5 times speedup (20\%) of 25--30\% comes out to 5--6\% saving in the total simulation time.}

This clarification has been added.

\item \textit{Line 92: “a” appears twice.}

Corrected -- moved to \textbf{box 2}.

\item \textit{Lines 161 and 163: is there a citation for CUBLAS and BLAS? If these are acronyms, what do they stand for?}

The acronyms has been defined in the revised text. References for CUBLAS and BLAS have been added.

\item \textit{Line 162: Saad, 2003 is misspelled.}

Corrected

\item \textit{Line 173: what is GEMV? Is this an acronym?}

It is an abbreviation for general matrix-vector (GEMV) product. This has been added to the manuscript.

\item \textit{Lines 193 and 196: Xeon is misspelled.}

Corrected

\item \textit{Lines 251--252: is this treatment of RCLOSE applied to PCG solver as well? If not, then the comparison between PCG and UPCG is unfair since the two solvers would converge to different RCLOSE values.}

Yes. This has been added.

\item \textit{Line 317: if MIC is more efficient, was parallelization of the MIC preconditioner attempted? Please make a statement as to why not.}

We did not intend to say anything more than the standard PCG solver is generally fairly efficient rather than something specific about the efficiency of the MIC preconditioner relative to the Jacobi, GLSPOLY, or MILU0 preconditioners available with the UPCG solver. This sentence has been revised to clarify this point.

\item \textit{Table 3: several values in this table are less than 1, indicating that UPCG is slower than PCG. What is the possible reason of this slowdown?}

Additional clarification has been provided for PCG(MIC)/UPCG-OpenMP speedups less than 1 in the \textbf{Discussion} section.

### 2012-09-04 ResponseTo_GW20120518-0099_Review.tex

\noindent
\textbf{Use of general purpose graphical processing units with MODFLOW-2005 \\ (Ground Water Manuscript GW20120518-0099) }\\

\noindent
\textbf{Authors:} Hughes, J.D., and White, J.T \\

\today \\

\noindent
We greatly appreciate the comments provided by the three reviewers and the editor. We have addressed all of the comments in this document. In general, we have made all of the suggested modifications. In a few cases we disagreed with reviewer comments but have provided a justification for not accepting the comment.

Reviewer comments are included (\textit{in italics}) prior to our responses. Revised material in the manuscript is indicated using \color{blue} blue \color{black}text. Text indicated in \color{cyan} cyan \color{black} in the revised manuscript represent changes in response to internal USGS reviewers.

@@

\end{enumerate}

\section*{Reviewer 2 Comments:}
\begin{enumerate}
\item \textit{Overall, lots of jargon is used. Naturally, this is an emerging field, and one that many readers of Ground Water will not be familiar with, so the use of jargon will be difficult for the uninitiated reader. For example, lines 75-79 are filled with terms that will mean nothing to the average groundwater modeler - consider deleting (or moving to appendix or online material) this paragraph and other similar paragraphs that might not hold much meaning to the average reader.}

We have tried to eliminate all non-essential GPGPU jargon and make the main text more readable for the typical Ground Water reader.  A significant portion of the jargon has been moved to Boxes 1 and 2 as suggested by the editor.

\item \label{kpropertydiscussion} \textit{Thorough comparisons were made for different preconditioners and different parallelization strategies and the authors did a great job of making the comparisons fair. Considering the results of the various preconditioners, it appears that Jacobi preconditioning is good enough for many of the problems, which leads me to believe that the K fields are fairly smooth (and the spread of the eigenvalues are nicely clustered). I tend to think of this as the exception in rather than the rule in most practical groundwater models. Can you provide guidance for using MIC vs. POLY vs. Jacobi based on the K field? In other words, I think these results are highly dependent on the K field and only 2 fields are used. How would the results vary if you stretched the contrasts in your K field? In that regard, to be fair, the GMG solver should also be used as a basis for comparison and give insight about which strategy should be used for optimal performance.}

Evaluation of the maximum and minimum eigenvalues for the original 1000 $\times$ 1000 problem suggests that there is clustering of eigenvalues in the original $\mathbf{A}$ matrix. It was not possible to evaluate the full spread of the eigenvalues because of the size of the $\mathbf{A}$ matrix. Evaluation of the eigenvalues of the final preconditioned matrix for the Jacobi preconditioner, however, indicates there are two clusters with the first ranging from approximately 2.5$\times 10^{-4}$ to 1$\times 10^{-2}$ and approximately 1.5$\times 10^{-2}$.

To look at the effect that eigenvalue clustering has on the effectiveness of the Jacobi preconditioner we have evaluated a number of alternative problems in an effort to reduce the smoothness of the fields. The additional fields that were evaluated are based on the fields used to evaluate the performance of the UPCG solver presented in the manuscript and included: (1) random permutations values in the original fields; (2) fields that used the average hydraulic conductivity in select quantiles of the original fields (to develop a zoned field); (3) a constant field based on the mean hydraulic conductivity of the original fields; and (4) a stochastic field that used the same properties of the original fields except with a variance that varied over 5 orders of magnitude (the original varied over 3 orders of magnitude).  Evaluation of the eigenvalues of the final preconditioned matrix for the Jacobi preconditioner indicates that the eigenvalues for (1) are clustered in three groups with a cluster covering a large range from approximately 5$\times 10^{-4}$ to 7.5$\times 10^{-3}$, (2) show no visible signs of clustering, (3) show no visible signs of clustering, and (4) are clustered into two groups like the original field but span a larger range.

For all of these alternative fields, the Jacobi preconditioner was able to sufficient to solve the problem and required a comparable number of iterations to solve the problem (approximately 4,500 iterations for the 1,000 $\times$ 1,000 $\times$ 1 problem) except for the fields with a larger variance (approximately 5,200 iterations for the 1,000 $\times$ 1,000 $\times$ 1 problem). The authors suspect the Jacobi preconditioner is sufficient in almost every case evaluated because of the number of iterations that are allowed (50,000). In fact, we were able to get the 4,000 $\times$ 4,000 $\times$ 1 problem to solve if the maximum number of iterations was increased (a solution was achieved in XX,XXX iterations).

We have not included results for any of these additional simulations but have added additional discussion of how speedup may be affected by the hydraulic conductivity field has been added to the \textbf{Discussion} and \textbf{Conclusions} sections.

@@

\item \textit{There seems to be a tradeoff between parallelizing on the CPU vs. the GPU depending on the sophistication of the preconditioner. What is not clear was if it Open-MP was used on PCG2. Reference is made to the work by to Dong and Li, but it appears that this was not considered. How do these results compare to that of Dong and Li? The intent here is not to say that your results are better/worse than those of Dong and Li. Simply, not everyone has access to a video card with double precision GPU capabilities, whereas multicore CPU's seem to be more common, so some guidance of which strategy might be more effective depending on the user's hardware would be helpful (ie, If I were buying a new workstation, should I focus on getting the best multicore CPU, or should I get a less powerful CPU and spend money on a high-end GPU?) I understand that the answer to this question is not straightforward, but any additional insights you could provide regarding this tradeoff would be helpful.}

OpenMP was not evaluated on PCG2 but was considered using the UPCG solver. We believe we can accurately report on conditions under which use of a GPGPU solution would be preferred to an OpenMP solution and vice-versa based on (1) the similarity of the effectiveness of the UPCG solver with the MILU0 preconditioner and the PCG solver with the MIC preconditioner and (2) the fact that UPCG solver speedups with OpenMP are comparable to the Intel compiler results reported in Dong and Li (2009) . We have revised the \textbf{Conclusions} section (see response to \textbf{Reviewer 2 Comment~\ref{kpropertydiscussion}}). We have also evaluated the effect that Intel compiler optimization has on OpenMP speedups and have revised the \textbf{Discussion} section to reflect the observation that speedups were comparable with and without use of the -O3 compiler switch.

We have also included our assessment of the best approach for being able to run a range of MODFLOW problems in parallel has been added to the \textbf{Conclusions} section.

@@

\item \textit{Line 351 - "developed to *use* the PCG" (one does not "solve" the algorithm) }

Suggested modification made.

\item \textit{Line 376 - as indicated above, can you provide some guidance here about when one approach might be better than another depending on the system properties (size, K field). Also, GMG should be included here to so that the full suite of approaches are compared (weak but highly parallelizable to strong but not very parallelizable). }

See response to \textbf{Reviewer 2 Comment~\ref{kpropertydiscussion}}.

@@

\item \textit{Figure 4 - I think this figure is extremely important as it shows where the speed ups occur. However, it is not clear how much time is spent in the GPU for the matrix-vector and BLAS operations because the bars seem to not show up on the scale? Can you include a value here so the reader does not assume these are zero. }

Figure 4 has been modified as suggested.

\item \textit{In light of the speed ups coming from CUBLAS operations being parallelized over the cells in the model, I am wondering if additional speed up could be obtained by using other primitives in the CUSparse library for the triangular solves and/or the decomposition? Was this considered as well? There is some new work that tries to parallelize ILU (see: http://research.nvidia.com/publication/parallel-incomplete-lu-and-cholesky-factorization-preconditioned-iterative-methods-gpu) }

We are aware of this work but have not attempted to implement it. One of the reasons we did not devote any time to its implementation is Naumov (2011) only reported an average speedup of 2.0 which is less than was observed with the GLSPOLY preconditioner. The work of Li and Saad (2010) pointed us in the direction of the GLSPOLY preconditioner. However, we have included a few sentences discussing recent work devoted to improving the effectiveness of incomplete factorization preconditioners in the \textbf{Conclusions} section.

@@

\item \textit{Tables should be reworked. I do not think we need 3 digits after decimal point for these comparisons. Please also consider significant digits in the tables.}

Double-precision floating point variables are used in the solvers and the tables have been rounded to single-precision numbers. Since we are reporting results from a numerical analysis we thought it appropriate to retain a larger number of digits to allow readers the ability to evaluate our results differently than we have. For example it would be possible for readers to calculate accurate run times (in seconds) for each of the simulations summarized in \textbf{table 2} using the speedup values and MIC cpu times. This may be of benefit for future studies conducted by other researchers. However, we would be willing to modify the number of digits reported if the editor agrees with this comment.

\item \textit{The authors only consider total execution time which includes time required for both Picard and linear solver iterations. Did you closely examined these iteration performance in solver comparisons? How these iteration performance changes among different preconditioning schemes? }

A table has been added (revised \textbf{Table 1}) which includes information on the number of iterations required for each of the solvers and various preconditioners. Additional discussion has been added in the \textbf{Discussion} section to provide some guidance on the strength of the preconditioners for the problems evaluated.

### 2012-09-04 ResponseTo_GW20120518-0099_Review.tex

Evaluation of the maximum and minimum eigenvalues for the original 1000 $\times$ 1000 problem suggests that there is clustering of eigenvalues in the original $\mathbf{A}$ matrix. It was not possible to evaluate the full spread of the eigenvalues because of the size of the $\mathbf{A}$ matrix. Evaluation of the eigenvalues of the final preconditioned matrix for the Jacobi preconditioner, however, indicates there are two clusters with the first ranging from approximately 2.5$\times 10^{-4}$ to 1$\times 10^{-2}$ and approximately 1.5$\times 10^{-2}$.

To look at the effect that eigenvalue clustering has on the effectiveness of the Jacobi preconditioner we have evaluated a number of alternative problems in an effort to reduce the smoothness of the fields. The additional fields that were evaluated are based on the fields used to evaluate the performance of the UPCG solver presented in the manuscript and included: (1) random permutations values in the original fields; (2) fields that used the average hydraulic conductivity in select quantiles of the original fields (to develop a zoned field); (3) a constant field based on the mean hydraulic conductivity of the original fields; and (4) a stochastic field that used the same properties of the original fields except with a variance that varied over 5 orders of magnitude (the original varied over 3 orders of magnitude).  Evaluation of the eigenvalues of the final preconditioned matrix for the Jacobi preconditioner indicates that the eigenvalues for (1) are clustered in three groups with a cluster covering a large range from approximately 5$\times 10^{-4}$ to 7.5$\times 10^{-3}$, (2) show no visible signs of clustering, (3) show no visible signs of clustering, and (4) are clustered into two groups like the original field but span a larger range.

For all of these alternative fields, the Jacobi preconditioner was able to solve the problem and required a comparable number of iterations (approximately 4,500 iterations for the 1,000 $\times$ 1,000 $\times$ 1 problem) except for the fields with a larger variance (approximately 5,200 iterations for the 1,000 $\times$ 1,000 $\times$ 1 problem). The authors suspect the Jacobi preconditioner is sufficient in almost every case evaluated because of the number of iterations that are allowed (50,000). In fact, we were able to get the 4,000 $\times$ 4,000 $\times$ 1 problem to solve if the maximum number of iterations was increased (a solution was achieved in XX,XXX iterations).

### 2012-09-04 GW20120518-0099-r01.tex
jwhite review

Parallelization of linear solvers is another approach for accommodating larger model sizes. \color{blue}Ideal parallelization would result in a linear speedup  (CPU runtime / parallel runtime) as additional CPUs and/or computers were applied to any given problem. For example, if ideal parallelization were achieved and 2, 3, 4, and 5 CPUs were applied to a problem the speedup would be 2, 3, 4, and 5, respectively. However, ideal parallelization is rarely, if ever, achieved because of communication bottlenecks which may be a result of hardware limitations and/or use of algorithms that are difficult to parallelize. \color{cyan}In this study, speedup is calculated using the amount of time spent in solver routines not the total runtime (which includes the time spent in other non-solver routines).\color{black}

@@

All of the models simulated steady-state conditions. As a result, specific storage and specific yield values were not specified for the aquifer. Initial heads were specified to be 0.0 m throughout the model domain, sufficiently different from final steady-state heads. Simulated heads \color{cyan} that satisfy specified convergence criteria \color{black} for the one layer 1,000 $\times$ 1,000 cell model are shown in \textbf{figure~\ref{FigModelDomain}} and show the effect that groundwater withdrawals and heterogeneous hydraulic conductivity have on water levels. \color{cyan}Simulated water-levels for simulations that satisfy specified convergence criteria and use different (1) solvers, (2) preconditioners, and/or (3) hardware are within $\pm$RCLOSE m.\color{black}

>>>>

Parallelization of linear solvers is another approach for accommodating larger model sizes. \color{blue}Ideal parallelization would result in a linear speedup  (CPU runtime / parallel runtime) as additional CPUs and/or computers were applied to any given problem. For example, if ideal parallelization were achieved and 2, 3, 4, and 5 CPUs were applied to a problem the speedup would be 2, 3, 4, and 5, respectively. However, ideal parallelization is rarely, if ever, achieved because of communication overhead which may be a result of hardware limitations and/or use of algorithms that are difficult to parallelize. \color{cyan}In this study, speedup is calculated using the amount of time spent in solver routines not the total runtime (which includes the time spent in other non-solver routines).\color{black}

@@

All of the models simulated steady-state conditions. As a result, specific storage and specific yield values were not specified for the aquifer. Initial heads were specified to be 0.0 m throughout the model domain, sufficiently different from final steady-state heads. Simulated heads \color{cyan} that satisfy specified convergence criteria \color{black} for the one layer 1,000 $\times$ 1,000 cell model are shown in \textbf{figure~\ref{FigModelDomain}} and show the effect that groundwater withdrawals and heterogeneous hydraulic conductivity have on water levels. \color{cyan}Water-level distributions for all converged simulations, including simulations using different (1) solvers, (2) preconditioners, and/or (3) hardware, are identical within the tolerance of $\pm$RCLOSE m.\color{black}
